### 2023-03-20 paper-thesis/chapters/21-adding-inference.tex
Improve more text

There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others. One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable. There are some situations in which this approach fails, but in most real-world scenarios it works perfectly. For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:

@@

A more powerful algorithm that could solve this is \emph{higher-order pattern unification} as defined by Miller~\cite{Miller89}, which is implemented in Agda. Implementing this in Statix is theoretically possible\footnote{Since Statix is turing-complete}, but would be quite difficult to do in practice.

@@

We would like to avoid implementing an algorithm at all, instead using Statix' built-in first-order unification to do the type inference for us. Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code (since Statix is not a general-purpose programming language), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.

However, we cannot immediately use Statix' built-in first-order unification (which acts in the meta language, Statix) to implement first-order unification in the object language. Ideally when implementing beta equality we would match on $\beq{e_1}{e_2}$ where $e_1$ is a free variable, but Statix does not allow for querying whether variables are free.

@@

\section{Implementing AFOU} \todo{Do we want all these examples in Statix or math notation?}
\label{implementing-inference}

First, we introduce a new constructor \verb|Infer : Expr -> Expr|, which denotes the variables which could be free. The constructor is introduced when we encounter a \verb|_| variable, a marker that something needs to be inferred.
\begin{lstlisting}
typeOfExpr(s, Var(Syn("_"))) = (Infer(q), qt) :-
	(_, qt) == typeOfExpr(sEmpty(), q).
\end{lstlisting}

Note that the type of \verb|typeOfExpr| has changed, it now returns two expressions, the first being the same expression that was passed in except with \verb|_| variables replaced with \verb|Infer| constructors, and the second being the type.
\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr * Expr
\end{lstlisting}

@@

The only difference between AFOU and FOU is case \ref{inf-case2} (for functions) and \ref{inf-case4}.

In case \ref{inf-case2}, this will only infer the value if it has not been inferred yet. This means that if the same value is inferred from two different places, this will infer two different values and fail.

@@

\todo{Mention Equational Unification as a solution}

>>>>

There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others. One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable. There are some situations in which this approach fails to infer a term, but in most real-world scenarios it works very well. For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:

@@

A more powerful algorithm that could solve this is \emph{higher-order pattern unification} as defined by Miller~\cite{Miller89}, which is implemented in Agda~\cite{agda}. Implementing this in Statix is theoretically possible\footnote{Since Statix is turing-complete}, but would be quite difficult to do in practice.

@@

We would like to avoid implementing an algorithm at all, instead using Statix' built-in first-order unification to do the type inference for us. Implementing an inference algorithm in Statix is theoretically possible but this would not be a clean implementation (since Statix is a domain specific language that is not meant to implement these algorithms), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.

However, we cannot immediately use Statix' built-in first-order unification (which acts in the meta language, Statix) to implement first-order unification in the object language. Ideally when implementing beta equality when we encounter $\beq{e_1}{e_2}$ we would check if $e_1$ is a free variable, but Statix does not allow for querying whether variables are free.

@@

\section{Implementing AFOU}
\label{implementing-inference}

First, we introduce a new constructor \verb|Infer : Expr -> Expr|, which denotes the variables which could be free. The constructor is introduced when we encounter a placeholder.
\begin{lstlisting}
typeOfExpr(s, Var(Syn("_"))) = (Infer(q), qt) :-
	(_, qt) == typeOfExpr(sEmpty(), q).
\end{lstlisting}

Note that the type of \verb|typeOfExpr| has changed, it now returns two expressions, the first being the same expression that was passed in except with placeholders replaced with \verb|Infer| constructors, and the second being the type.
\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr * Expr
\end{lstlisting}

@@

The only difference between AFOU and FOU is case \ref{inf-case2} (for functions) and case \ref{inf-case4}.

In case \ref{inf-case2}, we force the name of the function constructors to be equal. Ideally, we would generate a fresh name for the variable if and only if no name has been generated for it yet, but there is no way to tell whether this is the case so we use this approximation.

@@

We could solve this problem in two ways:

\begin{enumerate}
	\item We could always beta-reduce terms completely to their normal form, then terms that are beta-equal are also identical up to alpha equality. This would solve the problem in case \ref{inf-case4}, but the problem of case \ref{inf-case2} would remain. Furthermore, doing all this beta reduction is terrible for the performance of the type checker: Beta-reducing terms may take an arbitrary amount of time, though it is guaranteed to terminate since the Calculus of Constructions is strongly normalizing~\cite{Coquand_Huet_1988}.
	\item Ideally, we want a declarative definition of the problem. To do this, we need to be able to declare that two terms that are not identical are equal, which is possible under \emph{equational unification}~\cite[Section 2.1]{Siekmann86}. If this system was added to Statix, this would make implementing more powerful inference algorithms cleanly possible.
\end{enumerate}

### 2023-03-20 paper-thesis/chapters/22-adding-datatypes.tex
Improve more text

Another useful feature is support for inductive data types. An example of a simple inductive datatype is the \verb|Nat| type:
\begin{lstlisting}
data Nat : -> Type where
    zero : Nat,
    suc : Nat -> Nat;
\end{lstlisting}
We will design data types to have the same features as in Agda:
\begin{enumerate}
\item Data types may be recursive, that is the data type may take itself as one of the constructor arguments. This can be seen in the definition of \verb|Nat| above.
\item Parameters, which are datatypes that are polymorphic, and are required to be the same for all constructors, such as
\begin{lstlisting}
data Maybe (T : Type) : -> Type where
    None : Maybe T,
    Some : T -> Maybe T;
\end{lstlisting}
\item Indices, which are datatypes that are polymorphic, that may vary from constructor to constructor
\begin{lstlisting}
data Eq : (e1: Bool) (e2 : Bool) -> Type where
    refl : e : Bool -> Eq e e;
\end{lstlisting}
\item We can also combine parameters and indices, for example to create the generic \verb|Eq| type. Note that parameters and indices may also depend on earlier parameters and indices.
\begin{lstlisting}
data Eq (T : Type) : (e1 : T) (e2 : T) -> Type where
    refl : e : T -> Eq e e;
\end{lstlisting}
\item Data types must be strictly positive. This ensures that we cannot make non-terminating programs\footnote{After we add support for universes in chapter~\ref{ch:universes}}. For example, the following data type is forbidden, because it refers to itself in a negative position. We will explain exactly how this check works in section \ref{ch:datatypes:positivity}.
\begin{lstlisting}
data Bad : -> Type where
	bad : (Bad -> Bool) -> Bad;
\end{lstlisting}
\end{enumerate}

A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators~\cite{eliminators}. An eliminator is a function that can be used to case match on a data type. For example, the eliminator of the \verb|Nat| type above is:
\begin{lstlisting}
elim Nat : P: (v: Nat -> Type) -> P Z 
	-> (x: Nat -> P x -> P (S x)) -> n: Nat -> P n
\end{lstlisting}

The general type of an eliminator for a data type \verb|N| is:
\begin{lstlisting}
parameters 
-> P: (indices -> N params indices -> Type) 
-> constructors
-> indices
-> v: N params indices
-> P indices v
\end{lstlisting}
The meaning of all the arguments is:
\begin{itemize}
	\item \verb|parameters| are the parameters of the data type we want to eliminate, since these are constant among all constructors those can be at the start.
	\item \verb|P| is the type that this eliminator will return. It may depend on the value \verb|v| that is eliminated.
	\item \verb|constructors| is a function that eliminates each of the constructors of the datatype. To eliminate a constructor of the form \verb|C : args -> N params indices| it generates a function \verb|args -> P indices (C pars args)|. For recursive datatypes, a previous case is generated, such as the \verb|P x| in the eliminator of \verb|Nat| above.
	\item \verb|indices| are the indices of the value that is to be eliminated.
	\item \verb|v| is the value that is to be eliminated.
	\item Finally, \verb|P indices v| is the result of the elimination.
\end{itemize}

\section{Type-checking Data Type Declarations}
\label{ch:datatypes:typechecking}

The definition of \verb|type_check| for an inductive data type \verb|N| is, conceptually:
\begin{enumerate}
	\item Create a new scope \verb|s1| whose parent is the scope the declaration was in \verb|s0|.
	\item \label{name-decl}Declare \verb|N : Params -> Indices -> Type| in \verb|s1|.
	\item \label{datatype-decl}We use a new scope-graph relation \verb|datatype : ID -> Expr| and declare \\ \verb|N : DataType(name, params, indices, constructors)| in \verb|s1| so that we can access information about the data type later.

@@

	\item Create a chain of scopes starting in \verb|s2| that will contain a scope for each index. We will type-check each next index using the previous index' scope, so that one index can depend on parameters and previous indices. Call the end of this chain \verb|s3|.

@@

	\item Type-check the rest of the program with scope \verb|s4|.
\end{enumerate}
For example, for the following inductive datatype:
\begin{lstlisting}
data Eq (T : Type) : (e1 : T) (e2 : T) -> Type where
	refl : e : T -> Eq e e;
...
\end{lstlisting}
Figure \ref{scope-graph-eq} shows the scope graph generated by this data type.
\begin{figure}[!ht]
	\begin{framed}
		\centering
		\begin{tikzpicture}[scopegraph, node distance = 6.5em and 5em]
			% S0 and S1 node
			\node[scope] (s0) {$s0$};
			\node[scope, below = 2em of s0] (s1) {$s1$};
			\draw (s1) edge[lbl={P}] (s0);

@@

The type of an eliminator was discussed in section \ref{ch:datatypes:intro}. When we type check an \verb|elim N| expression, the following happens:
\begin{enumerate}
	\item Query the scope to find the data type declaration that we created in point \ref{datatype-decl} of section \ref{ch:datatypes:typechecking}. This gives us access to the parameters, indices and constructors of the data type that is being eliminated. 
	\item Using some tedious logic, create the type discussed in section \ref{ch:datatypes:intro}. This ends up being around 100 lines of Statix code.
\end{enumerate}

@@

@@

\begin{lstlisting}
data Bad : Set where
	bad : (Bad -> Bad) -> Bad
	--     A       B       C
	-- A is in a negative position, B and C are OK
\end{lstlisting}

The implementation is simply an extra check during the type checking of data type declarations. It checks that the datatype only refers to itself in positive positions.

We have now shown how to add inductive datatypes to the implementation. Next, we will Universes (chapter~\ref{ch:universes}) to the implementation.

>>>>

Another useful feature is support for inductive data types. Inductive data types declare types that are defined by a list of constructors that produce the type. An example of a simple inductive datatype is the \verb|Nat| type:
\begin{lstlisting}
data Nat : -> Type where
    zero : Nat,
    suc : Nat -> Nat;
\end{lstlisting}
The data type has two constructors: \verb|zero| represents the natural number zero, \verb|suc| takes a natural number and represents the successor of that number. For example, \verb|suc (suc zero)| represents 2.

We will design data types to have the same features as in Agda:
\begin{enumerate}
\item Data types may be recursive, that is the data type may take itself as one of the constructor arguments. This can be seen in the definition of \verb|Nat| above.
\item Data types may have \emph{parameters}, which are datatypes that are polymorphic over a certain value, that is required to be the same for all constructors, such as:
\begin{lstlisting}
data Maybe (T : Type) : -> Type where
    None : Maybe T,
    Some : T -> Maybe T;
\end{lstlisting}
\item Data types may have \emph{indices}, which are datatypes that are polymorphic over a certain value, that may vary from constructor to constructor, such as:
\begin{lstlisting}
data Eq : (e1: Bool) (e2 : Bool) -> Type where
    refl : e : Bool -> Eq e e;
\end{lstlisting}
\item We can also combine parameters and indices, for example to create the polymorphic \verb|Eq| type, which represents a proof that two values are equal. Note that the type of parameters and indices may depend on the value earlier parameters and indices. For example, \verb|e1| and \verb|e2| have type \verb|T|.
\begin{lstlisting}
data Eq (T : Type) : (e1 : T) (e2 : T) -> Type where
    refl : e : T -> Eq e e;
\end{lstlisting}
\item Data types must be strictly positive. This ensures that we cannot make non-terminating programs\footnote{After we add support for universes in chapter~\ref{ch:universes}}. For example, the following data type is forbidden, because it refers to itself in a negative position. We will explain exactly how this check works in section \ref{ch:datatypes:positivity}.
\begin{lstlisting}
data Bad : -> Type where
	bad : (Bad -> Bool) -> Bad;
\end{lstlisting}
\end{enumerate}

A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators as described in Inductive families~\cite{eliminators}. An eliminator is a function that can be used to case match on a data type. For example, the eliminator of the \verb|Nat| type above is:
\begin{lstlisting}
elim Nat : P: (v: Nat -> Type) -> P Z 
	-> (x: Nat -> P x -> P (S x)) -> n: Nat -> P n
\end{lstlisting}

The general type of an eliminator for a data type \verb|N| is:
\begin{lstlisting}
parameters 
-> P: (indices -> v: N params indices -> Type) 
-> constructors
-> indices
-> v: N params indices
-> P indices v
\end{lstlisting}
The meaning of all the arguments is:
\begin{itemize}
	\item \verb|parameters| are the parameters of the data type we want to eliminate, since these are constant among all constructors those should be specified at the start.
	\item \verb|P| is the type that this eliminator will return. It may depend on the value \verb|v| that is eliminated.
	\item \verb|constructors| is a function that eliminates each of the constructors of the datatype. To eliminate a constructor of the form \verb|C : args -> N params indices| it generates a function \verb|args -> P indices (C pars args)|. For recursive datatypes, a previous case is generated, such as the \verb|P x| in the eliminator of \verb|Nat| above.
	\item \verb|indices| are the indices of the value that is to be eliminated.
	\item \verb|v| is the value that is to be eliminated.
	\item Finally, \verb|P indices v| is the result of the elimination.
\end{itemize}

\section{Type-checking Data Type Declarations}
\label{ch:datatypes:typechecking}
In this section we will explain how to type check data type declarations using Statix.
The definition of \verb|type_check| for an inductive data type \verb|N| is, conceptually:
\begin{enumerate}
	\item Create a new scope \verb|s1| whose parent is the scope the declaration was in \verb|s0|.
	\item \label{name-decl}Declare \verb|N : Params -> Indices -> Type| in \verb|s1|.
	\item \label{datatype-decl}We use a new scope-graph relation \verb|datatype : ID -> Expr| and declare \\ \verb|N : DataType(name, params, indices, constructors)| in \verb|s1| so that we can access information about the data type later.

@@

	\item Create a chain of scopes starting in \verb|s2| that will contain a scope for each index. We will type-check each next index using the previous index' scope, so that one index can depend on parameters and previous indices. Call the end of this chain \verb|s3|. The scope \verb|s3| is not used in the rest of this process, but creating this chain is required to check the types of the indices.

@@

	\item Type-check the rest of the program with scope \verb|s4|.
\end{enumerate}
For example, figure \ref{scope-graph-eq} shows the scopes that were created during the type checking of the following inductive datatype:
\begin{lstlisting}
data Eq (T : Type) : (e1 : T) (e2 : T) -> Type where
	refl : e : T -> Eq e e;
\end{lstlisting}
\begin{figure}[!ht]
	\begin{framed}
		\centering
		\begin{tikzpicture}[scopegraph, node distance = 6.5em and 5em]
			% S0 and S1 node
			\node[scope] (s0) {$s0$};
			\node[scope, below = 2em of s0] (s1) {$s1$};
			\draw (s1) edge[lbl={P}] (s0);

@@

This section discusses how to type check eliminators. The type of an eliminator was already discussed in section \ref{ch:datatypes:intro}. When we type check an \verb|elim N| expression, the following happens:
\begin{enumerate}
	\item Query the scope to find the data type declaration that we created in point \ref{datatype-decl} of section \ref{ch:datatypes:typechecking}. This gives us access to the parameters, indices and constructors of the data type that is being eliminated. 
	\item Using some tedious but conceptually simple Statix rules, create the type discussed in section \ref{ch:datatypes:intro}. This ends up being around 100 lines of Statix code.
\end{enumerate}

@@

This is all that needs to happen to define inductive data types. Next, we will add positivity checking.

@@

To show how this is can go wrong, consider the following example:
\begin{lstlisting}
data False : Set where

data Bad : Set where
	bad : (Bad → False) → Bad;

self-app : Bad → False
self-app (bad f) = f (bad f)

absurd : False
absurd = self-app (bad self-app)
\end{lstlisting}

Under the curry-howard correspondence, \verb|Bad| states that \verb|Bad| is true if and only if \verb|Bad| implies false. Since such a statement is inconsistent, we can use it to prove \verb|False|. To prevent such statements, we don't allow inductive datatypes to refer to themselves negatively, checking this is called \emph{positivity checking}. This solution is not complete (it rejects valid data types), but it is the solution most often used in practice, such as in Agda~\cite{agda}, Lean~\cite{lean} and Coq~\cite{coq}.

The exact condition that is checked is that if a constructor has a function as its argument, then the argument type (\verb|A| in the example below) can not refer to the declared data type.

\begin{lstlisting}
data Bad : Set where
	bad : (Bad -> Bad) -> Bad
	--     A       B       C
\end{lstlisting}

The implementation is simply an extra check during the type checking of data type declarations, which is straight-forward to implement in Statix.

We have now shown how to add inductive datatypes to the implementation. Next, we will add  Universes (chapter~\ref{ch:universes}) to the implementation.
